Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-01-31 paper.tex
Update abstract.

\begin{abstract}
The field of synthetic biology focuses on creating modular components which can generate complex and controllable synthetic biological systems, but the intrinsic noise of gene regulation can create undesired effects in these systems. This noise is often treated as a nuisance which should be removed, but theoretical noise-exploiting controllers have been shown to use noise to their advantage to control difficult control problems which would be difficult otherwise. In such work, a difficult control problem similar to Maxwells Demon was proposed and showed that noise exploiting controllers could send two identical systems to different stable points reliably by exploiting the natural noise of gene regulation. Those controllers relied upon systems with no time delay well known parameters but, in this work the performance of noise-exploiting controllers are measured under time delay, parameter perturbation and system scale.

\end{abstract}
Keywords: Stochastic control, optogenetic, synthetic biology

%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: Stochastic control, optogenetic, synthetic biology
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol

@@

Synthetic biology seeks to create modular\cite{Ng2019} and orthogonal\cite{Liu2018} components to sense and actuate\cite{Sheets2020} complex logical systems\cite{Groseclose2020}, and which are capable of performing a variety of advanced biological behaviors\cite{Shin2020}. New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}. Great advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. These types of digital-synthetic actuators allows for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.

@@

In \cite{May2021}, it was suprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry. More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric). Three effective controllers were found which could do this, a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell, and a probabilistic Model Predictive Feedback Controller (pMPC) that observed only the primary cell and integrated a probabilistic model to represent the non-stationary probability distribution of the secondary cells in the same environment.

@@

\section{Methods}
We begin with a five species model for the production of an arbitratry protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
This model for the production of protein uses a split T7 domain ($T7_N$,and $T7_C$) which reversibly dimerizes in the presence of light to create a T7 dimer. The T7 monomer is created at a constant rate and degrades linearly with the number of monomer. The split monomers of T7 can bind to create a T7 dimer (equation ...) . The T7 dimer can then bind to a gene which enables the production of protein with a leak term, and the protein can bind to a secondary self-activated promoter described by a Hill function for the creation of more protein. This auto-regulatory motif in protein production is responsible for creating non-linearity in the system that enables noise-exploiting control. The final step was the degredation of protein

>>>>

\begin{abstract}
Synthetic biology seeks to create modular components that generate complex and controllable biological behaviors, often at microscopic scales. In single-cell applications, where important regulatory RNA or proteins are present in small numbers, intrinsic fluctuations cause biochemical `noise' that frequently results in undesired effects, and this noise is usually treated as a nuisance that must be removed. However, recent theoretical analyses have shown that noise can sometimes be used to one's advantage when seeking to manipulate nonlinear regulatory systems that would be difficult or impossible to control in a fully deterministic setting. Specifically, it was shown that a noise-exploiting optogenetic controller reminiscent of Maxwell's Demon could systematically force multiple identical systems to reach different independently-assigned stable points, even for a few special cases where the model is inexact or if observations were incomplete. This work expands upon those analyses to explore in greater detail how control performance is affected by errors or approximations in the model or by temporal delays in the observations. We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.  Together, these findings suggest that noise-exploiting control should be possible in real experiments where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.

\end{abstract}
Keywords: Optogenetic control, Stochastic gene regulation, Synthetic biology

%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: Stochastic control, optogenetic, synthetic biology
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol

@@

Synthetic biology seeks to create modular\cite{Ng2019} and orthogonal\cite{Liu2018} components to sense and actuate\cite{Sheets2020} complex logical systems\cite{Groseclose2020}, and which are capable of performing a variety of advanced biological behaviors\cite{Shin2020}. 
New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}. Great advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. These types of digital-synthetic actuators allow for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.

@@

In \cite{May2021}, it was surprisingly found that, when noise, auto-regulation, and noise-exploiting feedback control were all present at once, the system could be controlled to bring two similar cells to two different arbitrary points simultaneously (Figure \ref{cartoons}A), but when any one of these components were missing the system could not break its symmetry. More specifically, this required that the symmetry between the protein probability distributions of each cell would not be not the same (or asymmetric). Three effective controllers were found which could do this, a Fully Aware Feedback Controller (FAC) that observed the protein counts of all observed cells, a Partially Aware Feedback Controller (PAC) that observed the protein counts of only one cell, and a probabilistic Model Predictive Feedback Controller (pMPC) that observed only the primary cell and integrated a probabilistic model to represent the non-stationary probability distribution of the secondary cells in the same environment.

@@

\section{Methods}
We begin with a five species model for the production of an arbitrary protein through the activation of light activatible T7 polymerase (figure 1B) motivated by Baumschlager, et al.
This model for the production of protein uses a split T7 domain ($T7_N$,and $T7_C$) which reversibly dimerizes in the presence of light to create a T7 dimer. The T7 monomer is created at a constant rate and degrades linearly with the number of monomer. The split monomers of T7 can bind to create a T7 dimer (equation ...) . The T7 dimer can then bind to a gene which enables the production of protein with a leak term, and the protein can bind to a secondary self-activated promoter described by a Hill function for the creation of more protein. This auto-regulatory motif in protein production is responsible for creating non-linearity in the system that enables noise-exploiting control. The final step was the degredation of protein
